---
title: "Introduction to patliR: from a compound list to a ranked candidate table"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to patliR: from a compound list to a ranked candidate table}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(patliR)
```

This vignette runs the closed loop of patliR end to end -- a compound list to
a ranked candidate table and a report -- using only the small example dataset
bundled with the package (`system.file("extdata", ..., package = "patliR")`),
so every chunk below runs offline, with no internet access and no optional
Bioconductor/STRINGdb dependency. It is not a real analysis (six compounds,
two conditions) -- it exists to show the shape of the pipeline and let you
confirm your own install works before pointing it at real data.

Every step below is a pure transformation: `proj <- some_step(proj, ...)`.
`proj` is an immutable S4 object, but the durable source of truth is always a
plain CSV written to the project directory -- `patliR_load()` can rebuild the
whole project from those CSVs alone, in a fresh R session or on another
machine. See `vignette(package = "patliR")`'s companion documents --
`METHODS.md` (what every function computes, with citations) and `DESIGN.md`
(the cross-cutting design decisions) -- for the full detail this vignette
only summarizes.

## 1. Get compounds in

```{r}
proj <- patliR_project(tempfile("patliR_vignette_"))

compound_list <- read.csv(
  system.file("extdata", "input_compound_list.csv", package = "patliR")
)
proj <- prep_compounds(proj, compound_list, identifier = "pubchem")
compounds(proj)
```

`prep_compounds()` validates every SMILES with the CDK toolkit and assigns a
canonical form -- a compound whose structure does not parse is logged and
dropped, never silently kept.

A GC-MS abundance matrix (replicates x conditions) turns into a
presence/absence call per condition via `prep_binarize()`:

```{r}
abundance <- read.csv(
  system.file("extdata", "input_abundance_matrix.csv", package = "patliR"),
  check.names = FALSE
)
proj <- prep_binarize(proj, abundance)
binarizedMatrix(proj)
```

A compound is called present in a condition when its averaged abundance is
above that condition's own first quartile (Q1) of non-zero values -- a
simple detection threshold, not a statistical test; see `METHODS.md` for the
exact rule.

## 2. ADME and toxicity screening (fully local, no internet)

```{r}
proj <- adme_local(proj)
proj <- adme_filter(proj, rules = c("ro5", "veber", "ghose", "egan", "oprea"))
proj <- tox_local(proj, alert_sets = c("pains", "brenk"))

report <- tox_report(proj)
report$summary
```

None of `adme_local()`'s five rules, or `tox_local()`'s PAINS/Brenk
substructure alerts, is a verdict -- they flag a compound as physicochemically
atypical or structurally alert-prone, never as "bad". `tox_report()` always
attaches the same disclaimer for exactly this reason.

## 3. Bring in predicted targets and build the network

```{r}
proj <- targets_import_batch(
  proj,
  system.file("extdata", "import_targets", package = "patliR"),
  platform = "superpred"
)
proj <- network_build(proj)
patliRResults(proj, "network_edges")
```

`network_build()` constructs one bipartite compound-target graph per
condition -- compounds on one side, predicted targets on the other, an edge
where a prediction platform (here: SuperPred exports) says a compound likely
hits that target. Every later `network_*` function reads this graph; none of
them rebuild it.

## 4. Network analysis (offline pieces only)

```{r}
proj <- network_centrality(proj)
proj <- network_hub_penalty(proj)
proj <- network_module_robustness(proj, seed = 1)

patliRResults(proj, "network_centrality")
patliRResults(proj, "network_module_robustness")
```

`network_centrality()`'s `degree_norm`/`betweenness_norm` are comparable
between a compound and a target of the *same* condition (Borgatti & Everett,
1997) -- not across conditions of different size.
`network_module_robustness()` clusters the graph (Leiden by default) and
simulates removing its highest-degree nodes one at a time, reporting an
R-index of how much of the module stays connected.

A handful of `network_*` functions need external resources this vignette
deliberately does not call (real STRING/Open Targets/Bioconductor
dependencies), so they are shown for reference, not run here:

```{r, eval = FALSE}
# needs internet + STRINGdb
proj <- disease_genes_fetch(proj, disease = "type 2 diabetes mellitus")
disease_id <- unique(patliRResults(proj, "disease_genes")$disease_id)[1]
proj <- network_proximity(proj, disease = disease_id)
proj <- network_synergy(proj, disease = disease_id)

# needs clusterProfiler + org.Hs.eg.db (+ ReactomePA for db = "reactome")
proj <- network_enrich(proj, db = "go")
```

## 5. Rank candidates and generate a report

`rank_candidates()` needs the `RobustRankAggreg` package (a `Suggests`, not
required to install `patliR` itself) -- this chunk only runs if it is
present:

```{r, eval = requireNamespace("RobustRankAggreg", quietly = TRUE)}
proj <- rank_candidates(proj, condition = "FLO-ET")
patliRResults(proj, "rank_candidates")
```

`rank_candidates()` combines every criterion that has data (here: ADME pass
fraction and network centrality -- the two mandatory ones; module robustness
and hub penalty auto-detected since they were just computed) via Robust Rank
Aggregation (Kolde et al. 2012) -- comparing each criterion's *rank*, not its
raw value, so no single criterion's scale can dominate the combined score.
Every raw value, its per-criterion rank, and the final `rra_rank` are all
kept in the output table -- never collapsed to one opaque number.

```{r}
proj <- report_generate(proj, condition = "FLO-ET")
patliRResults(proj, "report_log")
```

`report_generate()` writes one self-contained HTML file summarizing every
step above for that condition -- open the path in `report_log` in a browser.
A step that was never run (e.g. `network_synergy()` in this vignette) shows
as a clearly labelled placeholder in the report, not a gap you have to
notice yourself.

## Where to go next

- `METHODS.md` -- what every function computes, its formula, and its
  citation, in plain language.
- `DESIGN.md` -- the package's cross-cutting design decisions (why an
  immutable project object, why CSV as the source of truth, when a
  `Suggests` dependency is added vs. hand-rolled).
- `ROADMAP.md` -- what is designed but not yet built.
- Every function's own help page (`?rank_candidates`, `?network_proximity`,
  ...) documents its exact arguments, return schema, and references.
